---
title: "QC of run1 and run2 (and run3)"
author: "Friederike Dündar"
date: "4/26/2021; updated `r Sys.Date()`"
output:
    html_document:
        toc: true
        code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE, cache.lazy = FALSE)
```

```{r libraries, chache=FALSE}
library(data.table)
library(magrittr)
library(ggplot2); theme_set(theme_bw(base_size = 16))
library(patchwork)

data_dir <- "~/Documents/Projects/2020-09_CZI_JamesJingli/Eric_scRNAseq/data/"
```


We currently have the following samples:

```{r results='asis'}
smptab <- fread(paste0(data_dir, "samplesEric.txt"))
smptab$Sample <- paste(smptab$tissue, smptab$surgery, smptab$timepoint, smptab$run, sep = "_")
ABCutilities::my_table(smptab)
```

For QC, I will:

* first combine all samples into one big SCE object (`paste0("sce_unfiltered_", Sys.Date(), ".rds")`)
* extract the colData with the QC metrics to
    - generate QC plots on my local machine and
    - determine the individual QC thresholds for each sample

Then I'll return to the server for filtering, batch correction, clustering, dimReds, and SingleR annotation.
I will combine samples of the same tissue.

## Reading data in

```{r loading_10X_output, eval=FALSE}
# on server:
#$ conda activate czi
options(menu.graphics=FALSE)
library(SingleCellExperiment)
library(DropletUtils)
library(scater)
library(scran)
library(magrittr)

data_dir <- "/scratchLocal/frd2007/2021-04_EricJames_CZI/data/"
smptab <- read.table(paste0(data_dir, "samplesEric.txt"), header=TRUE, stringsAsFactors = FALSE)
smptab$my_sample_name <- paste(smptab$tissue, smptab$surgery, smptab$timepoint, smptab$run, sep = "_")

smpls <- unique(smptab$sample)

scel <- lapply(smpls, function(x){
    mysmp <- subset(smptab, sample == x)$my_sample_name
    print(paste("Reading CellRanger output for", x))
    out.sce <- DropletUtils::read10xCounts(
        samples = paste0(data_dir, "CellRangerOutput/", x,
            "/outs/filtered_feature_bc_matrix"),
        sample.names = mysmp,
        version = "auto")
})

full_data <- do.call(cbind, lapply(scel, counts))
cell_info <- do.call(rbind, lapply(scel, colData))
gene_info <- rowData(scel[[1]])


## combine in one object
sce.all <- SingleCellExperiment(
    list(counts = full_data), 
    rowData = gene_info,
    colData = cell_info, 
    metadata = list(Samples = names(scel))
)

rm(scel); gc()

## remove completely uncovered genes
gnszero <- Matrix::rowSums(counts(sce.all)) == 0
sce.all <- sce.all[!gnszero, ]
#> dim(sce.all)
#[1]  27859 442307

## add cellnames
cellnames <- lapply(unique(sce.all$Sample), function(x){
    tmp <- sce.all[, sce.all$Sample==x]
    n_smp <- ncol(tmp)
    outnames <- paste(x, c(1:n_smp),sep=".")
    return(outnames)
}) %>% unlist
colnames(sce.all) <- cellnames

## add QC info
is.mito <- grepl("mt-", ignore.case = TRUE, rowData(sce.all)$Symbol)
sce.all <- addPerCellQC(sce.all, subsets=list(mitochondrial=is.mito))

mm <- lapply(unique(sce.all$Sample), function(x){
        ss <- sce.all[, sce.all$Sample == x]$subsets_mitochondrial_percent
        isOutlier(ss, type = "higher")})
sce.all$mito.discard <- unlist(mm)
sce.all$mito.discard <- ifelse(is.na(sce.all$mito.discard), FALSE, sce.all$mito.discard)

## determine GENE count thresholds for each Sample individually
gg <- lapply(unique(sce.all$Sample), function(x){
    ss <- log10(sce.all[, sce.all$Sample == x]$detected)
    isOutlier(ss)
})
sce.all$gene.discard <- unlist(gg)

## save colData
library(data.table)
cd <-colData(sce.all) 
cd$cell <- rownames(cd)
cd <- as.data.frame(cd) %>% as.data.table
saveRDS(cd, file = paste0("colData_unfiltered_", Sys.Date(), ".rds"))

saveRDS(sce.all, file = paste0("sce_unfiltered_", Sys.Date(), ".rds"))
```

## Finding filters

```{r local_data_load}
## load the colData locally for Rmd
qcall <- readRDS(paste0(data_dir, "colData_unfiltered_2021-06-15.rds"))
qcall$tissue <- gsub("_.*","", qcall$Sample)
qcall$surgery <- gsub(".*_([MIsham]+)_.*", "\\1", qcall$Sample)
qcall$day <- gsub(".*_(d[0-9]+)_.*", "\\1", qcall$Sample)
qcall$run <- gsub(".*_","",qcall$Sample)
```

```{r fig.width = 18, fig.height = 12}
for(i in unique(qcall$tissue)){
    dt <- qcall[tissue == i]
    P1 <- ggplot(dt,
        aes(x = Sample, y = subsets_mitochondrial_percent, color = mito.discard)) +
        ggbeeswarm::geom_quasirandom(size = .2, alpha = .3, shape = 1) +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=14)) +
        scale_color_manual(values = rev(c("firebrick2","grey75"))) +
        ylab("% mitochondrial genes") + coord_cartesian(ylim = c(0,20)) +
        theme(legend.position = "bottom")+
        guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3, shape = 19)))
    P2 <- ggplot(dt,
        aes(x = subsets_mitochondrial_percent, y = log10(detected),
            color = mito.discard)) +
        geom_point(size = .2, shape = 1 , alpha = .5) +
        facet_wrap(~Sample, ncol = 3) +
        scale_color_manual(values = rev(c("firebrick2","grey75"))) +
        xlab("% mitochondrial genes") + ylab("log10(total number of genes)")+ 
        theme(legend.position = "bottom")+
        guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3, shape = 19)))
    pw <- P1 | P2
    pw <- pw + plot_annotation(title = i) + plot_layout(widths = c(1.5,3))
    print(pw)
}
```

## First filtering

```{r}
qcall[, 
    max_mito := ifelse(tissue == "islets", 10, 
        ifelse(tissue == "blood" & Sample %in% c("blood_MI_d30_run1","blood_sham_d30_run1", "blood_sham_d07_run2", "blood_MI_d07_run2"), 7.5,
            ifelse(tissue == "blood" & Sample != "blood_MI_d07_run2", 5,
                ifelse(tissue == "blood", 2,
                    ifelse(tissue == "heart" & surgery == "MI", 10,
                        ifelse(tissue == "heart", 15,
                            ifelse(tissue == "liver", 50, NA)))))))]
```

```{r fig.width = 18, fig.height = 12}
for(i in unique(qcall$tissue)){
    dt <- qcall[tissue == i & subsets_mitochondrial_percent <= max_mito]
    ymax <- max(dt$max_mito)
    P1 <- ggplot(dt,
        aes(x = Sample, y = subsets_mitochondrial_percent, color = gene.discard)) +
        ggbeeswarm::geom_quasirandom(size = .2, alpha = .3, shape = 1) +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=14)) +
        scale_color_manual(values = rev(c("blue","grey75"))) +
        ylab("% mitochondrial genes") + coord_cartesian(ylim = c(0,ymax)) +
        theme(legend.position = "bottom") +
        guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3, shape = 19)))
    P2 <- ggplot(dt,
        aes(x = subsets_mitochondrial_percent, y = log10(detected),
            color = gene.discard)) +
        geom_point(size = .2, shape = 1 , alpha = .5) +
        facet_wrap(~Sample, ncol = 3) +
        scale_color_manual(values = rev(c("blue","grey75"))) +
        xlab("% mitochondrial genes") + ylab("log10(total number of genes)")+ 
        theme(legend.position = "bottom") +
        guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3, shape = 19)))
    pw <- P1 | P2
    pw <- pw + plot_annotation(title = i) + plot_layout(widths = c(1.5,3))
    print(pw)
}
```

## Second filtering

Here, I am setting blood aside for the moment until I know via SingleR what the proper filters might be.

```{r }
qcall[, max_mito := ifelse(tissue == "islets",7.5, 
    ifelse(tissue == "liver", 10,
        ifelse(tissue == "heart", 10, max_mito)))]

qcall[, min_genes := ifelse(tissue == "islets", 3.25,
        ifelse(tissue == "heart", 3,
            ifelse(tissue == "liver", 2.6, NA)))]

qcall[ , gene.discard := ifelse(is.na(min_genes), FALSE,
    ifelse(log10(detected) <= min_genes, TRUE, FALSE))]
qcall[ , mito.discard := ifelse(subsets_mitochondrial_percent > max_mito, TRUE, FALSE)]
qcall[, rm_cell := ifelse(mito.discard == TRUE | gene.discard == TRUE, TRUE, FALSE)]
```

How many cells are going to be removed per sample? [Note that blood still needs to be figured out]

```{r }
table(qcall$Sample, qcall$rm_cell)
```
```{r fig.width = 10, fig.height=10}
ggplot(qcall, aes(x = Sample, fill = rm_cell)) +geom_bar() + coord_flip()  +
  #  scale_color_manual(values = c("midnightblue","dodgerblue3","dodgerblue1")) +
    scale_fill_manual(values = c("grey75","orange")) +
    ylab("# cells") + ggtitle("How many cells will be removed per sample?") +
    facet_wrap(~run, scales="free_y", ncol = 1)
```


```{r fig.width = 18, fig.height = 12}
for(i in unique(qcall$tissue)){
    dt <- qcall[tissue == i & subsets_mitochondrial_percent <= max_mito]
    ymax <- max(dt$max_mito)
    P1 <- ggplot(dt,
        aes(x = Sample, y = subsets_mitochondrial_percent, color = rm_cell)) +
        ggbeeswarm::geom_quasirandom(size = .2, alpha = .3, shape = 1) +
        theme(axis.text.x  = element_text(angle=90, vjust=0.5, size=14)) +
        scale_color_manual(values = rev(c("orange","grey75"))) +
        ylab("% mitochondrial genes") + coord_cartesian(ylim = c(0,ymax)) +
        theme(legend.position = "bottom") +
        guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3, shape = 19)))
    P2 <- ggplot(dt,
        aes(x = subsets_mitochondrial_percent, y = log10(detected),
            color = rm_cell)) +
        geom_point(size = .2, shape = 1 , alpha = .5) +
        facet_wrap(~Sample, ncol = 3) +
        scale_color_manual(values = rev(c("orange","grey75"))) +
        xlab("% mitochondrial genes") + ylab("log10(total number of genes)")+ 
        theme(legend.position = "bottom") +
        guides(colour = guide_legend(override.aes = list(alpha = 1, size = 3, shape = 19)))
    pw <- P1 | P2
    pw <- pw + plot_annotation(title = i) + plot_layout(widths = c(1.5,3))
    print(pw)
}
```

## Do the filtering

```{r eval=FALSE}
# save the colData to be added back to the SCE
saveRDS(qcall, file = paste0(data_dir, "colData_filtered_2021-06-15.rds"))
#$ scp ~/Documents/Projects/2020-09_CZI_JamesJingli/Eric_scRNAseq/data/colData_filtered_2021-06-15.rds frd2007@redteam2.pbtech:/scratchLocal/frd2007/2021-04_EricJames_CZI/2021-06-03_secondRunQC/

## on server
qcall <- readRDS("colData_filtered_2021-06-15.rds")
qcall <- qcall[, c("Sample","Barcode","cell","tissue","surgery","day","run","rm_cell")] %>%
    .[cd, on = c("Sample", "Barcode","cell")]

qcall.df <- DataFrame(as.data.frame(qcall))
rownames(qcall.df) <- qcall.df$cell
colData(sce.all) <- qcall.df[colnames(sce.all),]

## remove cells
sce.all <- sce.all[, !sce.all$rm_cell]
gnszero <- Matrix::rowSums(counts(sce.all)) == 0
sce.all <- sce.all[!gnszero, ]

saveRDS(sce.all, file = "sce_filtered_2021-06-15.rds")
```

Next step: SingleR-based cell labeling (see `2021-06-07_SingleR.Rmd`)
